---
title: "RNA-seq Analysis in R"
subtitle: "Annotation and Visualisation of RNA-seq results - Solutions"
author: "Stephane Ballereau, Dominique-Laurent Couturier, Mark Dunning, Abbi Edwards, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
minutes: 300
layout: page
---

```{r setup, include=FALSE}
library(EnsDb.Mmusculus.v79)
library(DESeq2)
library(tidyverse)
```

```{r prepareData, echo=FALSE, message=FALSE, warning=FALSE}
load("../Robjects/DE.RData")
```


## Retrieve full annotation

> ### Challenge 1 {.challenge}

> A reminder of the code we ran:

```
# lets set it up
ourCols <- c("SYMBOL", "GENEID", "ENTREZID")
ourKeys <- rownames(resLvV)[1:1000]

# run the query
annot <- AnnotationDbi::select(EnsDb.Mmusculus.v79, 
                keys=ourKeys, 
                columns=ourCols, 
                keytype="GENEID")
```

> That was just 1000 genes. We need annotations for the entire results table.
>   
> (a) Run the same query using all of the genes in our results table (`resLvV`)  
>
> (b) Can we also have the biotype of our genes too? Hint: You can find the name of the column for this by running columns(EnsDb.Mmusculus.v79)
>
> (c) How many Ensembl genes have multipe Entrez IDs associated with them?


```{r solutionChallenge1code, eval=FALSE, message=FALSE, warnings=FALSE}

# lets set it up
ourCols <- c("SYMBOL", "GENEID", "ENTREZID", "GENEBIOTYPE")
ourKeys <- rownames(resLvV)

# run the query
annot <-AnnotationDbi::select(EnsDb.Mmusculus.v79, 
                keys=ourKeys, 
                columns=ourCols, 
                keytype="GENEID")
# multiple EntrezIDs
multiples <- annot %>%
  add_count(GENEID) %>%  
  dplyr::filter(n>1)

length(unique(multiples$SYMBOL))

```

> ### Challenge 2 {.challenge}

If you haven't already make sure you load in our data and annotation. Then shrink the values. You can copy and paste the code below.

```
# First load data and annotations
load("../Robjects/DE.RData")
load("../Robjects/Ensembl_annotations.RData")

#Shrink our values
ddsShrink <- lfcShrink(ddsObj, coef="Status_lactate_vs_virgin")
shrinkLvV <- as.data.frame(ddsShrink) %>%
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)
```

> Use the log2 fold change (`logFC`) on the x-axis, and use `-log10(pvalue)` on the y-axis.
> (This `-log10` transformation is commonly used for p-values as it means that more
> significant genes have a higher scale) 
>
> (a) Create a new column of -log10(pvalue) values in shrinkLvV
>
> (b) Create a plot with points coloured by pvalue < 0.05 similar to how we did in the MA plot


```{r volcanoPlot, fig.height=5, fig.width=7}
# First load data and annotations
load("../Robjects/DE.RData")
load("../Robjects/Ensembl_annotations.RData")

#Shrink our values
ddsShrink <- lfcShrink(ddsObj, coef="Status_lactate_vs_virgin")
shrinkLvV <- as.data.frame(ddsShrink) %>%
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)

# first remove the filtered genes (FDR=NA) and create a -log10(FDR) column
filtTab <- shrinkLvV %>% 
    filter(!is.na(FDR)) %>% 
    mutate(`-log10(FDR)` = -log10(FDR))

ggplot(filtTab, aes(x = logFC, y=`-log10(FDR)`)) + 
    geom_point(aes(colour=FDR < 0.05), size=1)
```


